Data Specialization for a Staged Lambda Calculus

### 2014-11-07 paper/splitting.tex
Trying harder to get a transition to stage 2 splitting.

Given a term $e$ in the source language \lang, where terms of
stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage~\bbone\ and stage~\bbtwo\
subcomputations into two terms, $e_1$ and $e_2$, in an unstaged target language \langmono\ 
%each performing the
%subcomputations corresponding to each stage.  More precisely, given a
%\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
%unstaged target language \langmono\
such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluation of~$e$ in
the staged language produces the same result as evaluating~$e_1$, followed by evaluating 
$e_2$ with the result from~$e_1$.

@@

%Like stage \bbtwo\ splitting, splitting a stage~\bbone\ term $\colone e A$ produces a term $c$ containing all stage~\bbone\ subcomputations in $e$ and a resumer function %$l.r$ containing all stage~\bbtwo\ subcomputations.  
%which respectively contain all the stage \bbone\ and stage \bbtwo\
%subcomputations of $e$. 
%But where the stage \bbtwo\ splitting goal was phrased terms ensure.

%The situation is more complex than when splitting stage
%\bbtwo\ terms; to see why, consider the correctness criterion for splitting
%stage \bbone\ terms.

%Stage \bbone\ evaluation of $e$, via $e\mathbin{\redonesym}[\xi;v]$,
%yields a table of (stage \bbtwo) residuals $\xi$, and a (stage \bbone) partial
%value $v$.

% Ergo, if we intend to compare the results of evaluation to the those of splitting, 
% then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.

% But the whole point of splitting is that its results (and their evaluated forms) are monostage.

% Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.

% For functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
% replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.

%The only exception is for
%functions $\lam xAe$; such a value represents a computation, so masking splits
%its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
%stage \bbone\ component, the combined result.

%They operate by replacing all stage \bbone\
%(resp., \bbtwo) components of a partial value by \texttt{()}.

%It is tempting to compare the two results of evaluating $e$ to the
%two results of splitting it.

%To see where this goes wrong, consider the stage
%\bbone\ term
%\begin{lstlisting}
%1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
%\end{lstlisting}
%which evaluates to
%\begin{lstlisting}
%2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
%\end{lstlisting}
%This partial value contains a $\next$, and so is itself multistaged:
%intuitively, its stage \bbone\ part is \texttt{3}, while its stage \bbtwo\ part
%is $\mathtt{\hat{y}}$, which is bound to \texttt{7+5} in the table.

%We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
%$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
%a partial value at each stage, essentially by replacing all stage \bbone\
%(resp., \bbtwo) components of a partial value by \texttt{()}.

% % Umut: I made a pass over this, see above.  I don't think I lost
% % anything, so it should be fine to delete this.
% % \section{Splitting Algorithm}
% % \label{sec:splitting}
% % Although terms in \lang\ clearly distinguish their stage~\bbone\ and stage
% % \bbtwo\ components, these components may be interleaved in \lang\ code. The
% % goal of \emph{stage splitting} is to statically separate the stage \bbone\ and \bbtwo\ subcomputations
% % of a term $e$ in \lang\ into a pair of
% % terms $e_1$ and $e_2$ in an unstaged language \langmono, where $e_1$ performs all the
% % stage~\bbone\ subcomputations in $e$ and $e_2$ performs all of the stage~\bbtwo\ subcomputations.
% % Informally, splitting must ensure that running $e_2$ on the output of $e_1$ yields the same result 
% % as evaluating $e$ in \lang. (We make this intuition precise in the subsequent subsections.)

% % Since the mechanics of splitting stage~\bbtwo\ terms are conceptually simpler than those for stage \bbone\
% % terms, we begin by describing how to split stage \bbtwo\ terms before proceeding to more complicated cases.

% % \subsection {Splitting Stage \bbtwo\ Terms}

% % Splitting a stage~\bbtwo\ term $\coltwo e A$ generates
% %  a \emph{precomputation} $p$, and a \emph{resumer} function $l.r$. Whereas direct evaluation of $e$ in \lang\ reduces the stage \bbone\ subcomputations
% % of $e$ in place to create a residual, splitting statically isolates these computations into
% % $p$.  The resumer accepts the value produced by the precomputation as input
% % % (bound to the free variable $l$)
% % and executes all stage~\bbtwo\ subcomputations.   
% % Specifically, if $e$ in \lang\ reduces to the final value $v$, 
% % then after splitting the term $\letin l p r$ should reduce to the same value in \langmono.
% % (Thus, the resumer, bound to the value output by $p$, is equivalent to the residual term defined in~\ref{ssec:dynamics})

% % %	\ur{Correctness will have
% % %	  to go beyond this, by saying for example that all stage 1 parts are
% % %	  in p etc.  Note that this correctness allows a trivial split
% % %	   that shoves everything to r, e.g., p = () and r is ``e''
% % %	  (might need to flatten).}

% % Splitting stage~\bbtwo\ terms, as implemented by the judgment $e \splittwosym
% % [p,l.r]$ in \cref{fig:termSplitTwo} is defined recursively on the structure of $e$.
% % In the terminal cases of constants and variables, splitting generates precomputations that are \texttt{()}
% % (these terms do not involve any stage~\bbone\ computations) and resumers that produce the original term's value.
% % For example, the integer constant \texttt{5} splits into the precomputation \texttt{()} and \texttt{l=>5}.

% % For all non-terminal stage~\bbtwo\ terms $e$ in \lang\ (except $\prev$ and $\pause$), splitting descends into $e$, recursively splitting subterms.  
% % The precompuation is formed by combining the resultant $n$ subterm precomputations into a tuple: $p=(p_1,\ldots,p_n)$.
% % The generated resumer has the same structure as $e$, with every subterm replaced by its resumer.
% % Notice that splitting \texttt{case}s and functions causes the precomputation of subterms to be lifted out from
% % underneath stage \bbtwo\ binders.  % TODO: should probably draw a parallel to the same behavior in dynamics

% % The $\prev$ and $\pause$ rules depends on
% % splitting stage \bbone\ terms, and so they are covered later.

% % %For example, the stage \bbtwo\ term
% % %\begin{lstlisting}
% % %2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
% % %\end{lstlisting}
% % %has two subterms, the latter of which itself has two subterms. Two are $\pause$s
% % %containing stage \bbone\ computations, while the third has no stage \bbone\
% % %content. 
% % %
% % %\ur{The discussion of the subterms is not helpful. what is the point?}
% % %
% % %The precomputation \emph{lifts} these computations in the same tree
% % %structure: \verb|(1+2,(3+4,()))|.%
% % %\ur{Not clear what a ``tree structure'' is. }
% % %\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
% % %The resumer is a function which computes the stage \bbtwo\ value of the original
% % %term by mapping the results of the precomputation in place of the $\pause$s:
% % %\verb|(L1,(L2,_)) => L1 < (L2 - 5)|. 
% % %
% % %\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}
% % %
% % %Non-staging constructs with $n$ subterms have precomputations which are
% % %$n$-tuples of their subterms' precomputations, and resumers which pass each
% % %component to the resumer of the respective subterm.

% % \subsection{Splitting Stage \bbone\ Terms}

% % The goal of stage 2 splitting was to generate a precomputation and resumer such that the resumer and the \lang\ residual reduce to the same final value. However, stage~\bbone\ terms do not reduce to final values, rather they reduce to residual tables and partial values.  For example, consider the stage 1 term: 
% % \begin{lstlisting}
% % 1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
% % \end{lstlisting}
% % which evaluates (via $\mathbin{\redonesym}$), to the following residual table and a multi-stage tuple:

% % \begin{lstlisting}
% % 2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
% % \end{lstlisting}

% % %Like stage \bbtwo\ splitting, splitting a stage~\bbone\ term $\colone e A$ produces a term $c$ containing all stage~\bbone\ subcomputations in $e$ and a resumer function %$l.r$ containing all stage~\bbtwo\ subcomputations.  
% % %which respectively contain all the stage \bbone\ and stage \bbtwo\
% % %subcomputations of $e$. 
% % %But where the stage \bbtwo\ splitting goal was phrased terms ensure.

% % %The situation is more complex than when splitting stage
% % %\bbtwo\ terms; to see why, consider the correctness criterion for splitting
% % %stage \bbone\ terms.

% % %Stage \bbone\ evaluation of $e$, via $e\mathbin{\redonesym}[\xi;v]$,
% % %yields a table of (stage \bbtwo) residuals $\xi$, and a (stage \bbone) partial
% % %value $v$.

% % Note that this result simultaneously represents information at {\em both} stages: 
% % a stage~\bbone\ value 3, and the stage~\bbtwo\ computation $\letin \yhat {\mathtt{7+5}} \yhat$. 
% % Therefore, to make the results of splitting match the the partial value above, we choose to split \bbone\ terms in \lang\ into a resumer that reduces to the same value as the stage~\bbtwo\ computation, and an $e1$ that not only performs the precomputation expected by the resumer, but also computes values that match the stage~\bbone\ aspects of the partial value.

% % % Ergo, if we intend to compare the results of evaluation to the those of splitting, 
% % % then we'll need some mechanism to project from the residual table and partial value only the parts that are relevant at a particular stage.

% % % But the whole point of splitting is that its results (and their evaluated forms) are monostage.

% % % Intuitively for splitting TO MATCH THIS THING ABOVE.  It needs to match the 3, and the resumer should reduce to the same value as the reified partial value.

% % \subsubsection{Partial Value Masking}

% % To concretize the meaning of ``match", we introduce two {\em partial value masking} functions, $\masko{\cdot}$ and $\maskt{\cdot}$, that project from the residual table and partial value only the subterms that are relevant at a particular stage.  The masking functions are defined in \ref{fig:valMask}.

% % Since \texttt{next}'s only contain references to the residual table (computations that occur in stage~\bbtwo),
% % the primary operation of $\masko{\xi;v}$ is to replace all $\next$s in $v$ with \texttt{()}.   The result is a \langmono\ value with the same structure as $v$, but containing only its stage~\bbone\ information. Masking is more complex when $v$ is a function since the function's body itself may contain stage~\bbtwo\ subterms.  As a result, masking of functions depends on the general behavior of stage~\bbone\ splitting, described shortly.

% % While $\masko{\cdot}$ produces a value in \langmono,
% % $\maskt{\xi;v}$ produces a \emph{term} in \langmono\ that corresponds to the stage~\bbtwo\ computation represented by $[\xi;v]$.
% % Stage~\bbone\ information in $v$ is discarded by replacing base constants with \texttt{()}, replacing $\next~\hat{y}$ with $\hat{y}$, and eliminating all injection tags. (The sum type is at stage \bbone, so the tag is not relevant at stage \bbtwo.)
% % Last, $\maskt{\cdot}$ reifies $\xi$ around the modified $v$ to produce a term in \langmono. 
% % As with $\masko{\cdot}$, function masking by $\maskt{\cdot}$ depends on general term splitting.

% % % For functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
% % % replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.

% % %The only exception is for
% % %functions $\lam xAe$; such a value represents a computation, so masking splits
% % %its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
% % %stage \bbone\ component, the combined result.

% % %They operate by replacing all stage \bbone\
% % %(resp., \bbtwo) components of a partial value by \texttt{()}.

% % %It is tempting to compare the two results of evaluating $e$ to the
% % %two results of splitting it.

% % %To see where this goes wrong, consider the stage
% % %\bbone\ term
% % %\begin{lstlisting}
% % %1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
% % %\end{lstlisting}
% % %which evaluates to
% % %\begin{lstlisting}
% % %2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
% % %\end{lstlisting}
% % %This partial value contains a $\next$, and so is itself multistaged:
% % %intuitively, its stage \bbone\ part is \texttt{3}, while its stage \bbtwo\ part
% % %is $\mathtt{\hat{y}}$, which is bound to \texttt{7+5} in the table.

% % %We formalize this notion via two \emph{masking} functions (\ref{fig:valMask}),
% % %$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
% % %a partial value at each stage, essentially by replacing all stage \bbone\
% % %(resp., \bbtwo) components of a partial value by \texttt{()}.

Splitting a stage \bbone\ term in \lang\
% $\colone e A$ 
(where $e\mathbin{\redonesym}[\xi;v]$), yields a term 
$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
To ensure the results of splitting produce the same output as the evaluation $e$ in \lang,
we require that $e_1$ reduce to the tuple $(y,z)$,
where $y$ is identical to $\masko{\xi;v}$, and 
$\letin l z r$ reduces to the same value as $\maskt{\xi;v}$. Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce an \emph{immediate result} $y$, and the stage~\bbone\ subcomputations (called \emph{precomputations}) required to produce the input required by the resumer, we refer to this term as the \emph{combined result}, and hereafter denote it as $c$.

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplitOne}, proceeds
recursively on the structure of~$e$.
When $e$ is a terminal (a variable or base constant)
splitting yields a combined result formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}. (Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, \texttt{3} splits into the combined result \texttt{(3,())} and resumer \texttt{\_=>()}.

@@

% We call attention to the rule for {\tt case}s since it, and the rule for $\pause$, where splitting adds non-trivial logic to the precomputation.

@@

Splitting a stage~\bbtwo\ term $\coltwo e A$ generates a
\emph{precomputation}~$p$, and a~\emph{resumer} function $l.r$.  The
resumer accepts the value produced by the precomputation as input
($l$) and performs the stage~\bbtwo\ subcomputations of~$e$.
Specifically, if~$e$ in~\lang\ reduces to the value~$v$, then after
splitting the term $\letin l p r$ also reduces to~$v$ in \langmono.

\ur{I don't quite understand the following.}
(Thus, the resumer, bound to the value output by $p$, is equivalent to
the residual term defined in~\ref{ssec:dynamics})

%	\ur{Correctness will have
%	  to go beyond this, by saying for example that all stage 1 parts are
%	  in p etc.  Note that this correctness allows a trivial split
%	   that shoves everything to r, e.g., p = () and r is ``e''
%	  (might need to flatten).}

The splitting algorithm for stage~\bbtwo\ terms, as specified by the
judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, proceeds
recursively on the structure of~$e$.  In the terminal cases of
constants and variables, splitting generates precomputations that are
\texttt{()}, because these terms do not involve any stage~\bbone\
computations, and resumers that produce the original term. For
example, the integer constant \texttt{5} splits into the
precomputation \texttt{()} and \texttt{l=>5}.

>>>>

Given a term $e$ in the source language \lang, where terms of
stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage~\bbone\ and stage~\bbtwo\
subcomputations into two terms, $e_1$ and $e_2$, in an unstaged target language \langmono\ 
%each performing the
%subcomputations corresponding to each stage.  More precisely, given a
%\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
%unstaged target language \langmono\
such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluating~$e$ in
the staged language produces the same result as evaluating~$e_1$ in the target language, then further evaluating 
$e_2$ in the target language with the result from~$e_1$.

@@

Splitting a stage \bbone\ term $e$ in \lang\
% $\colone e A$ 
yields a term 
$e_1$ that performs all stage~\bbone\ subcomputations in $e$ and a \emph{resumer} function $e_2 = l.r$ that performs all stage~\bbtwo\ subcomputations.
To ensure that the results of splitting produce the same output as the evaluation $e\mathbin{\redonesym}[\xi;v]$ in \lang,
we require that $e_1$ reduce to the tuple $(y,z)$,
where $y$ is identical to $\masko{\xi;v}$, and 
$\letin l z r$ reduces to the same value as $\maskt{\xi;v}$. Since $e_1$ performs both stage~\bbone\ subcomputations needed to produce $y$ (called the \emph{immediate result}), and the stage~\bbone\ subcomputations (called \emph{precomputations}) that produce the input required by the resumer, we refer to this term as the \emph{combined term}, and hereafter denote it as $c$.

The splitting algorithm for stage~\bbone\ terms, as specified by the
judgment $e \splitonesym [c,l.r]$ in \cref{fig:termSplitOne}, proceeds
recursively on the structure of~$e$.
When $e$ is a terminal (a variable or base constant)
splitting yields a combined term formed by tupling $e$ with a \texttt{()} precomputation, and the trivial resumer \texttt{()}. (Stage~\bbone\ terminals, by definition, contain no stage~\bbtwo\ content.)  For example, \texttt{3} splits into the combined term \texttt{(3,())} and resumer \texttt{\_=>()}.

@@

@@

Rather than match the

Splitting stage~\bbtwo\ terms in \lang\ takes on a simpler form than stage~\bbone\ terms since stage~\bbtwo\ terms because rather that the output of splitting

Like stage one, we get two terms.

But now the objective is to match the partial value, but the value after full evaluation.

That means we need to match values.

Splitting a stage~\bbtwo\ term $e$ in \lang\ generates a
precomputation~$p$ that performs all stage~\bbone\ subcomputations in $e$,
and a~resumer function $e_2=l.r$ that performs all stage~\bbtwo\ subcomputations.
If full evaluation of $e$ in~\lang\ reduces to the value~$v$, then $\letin l p r$ must reduce to~$v$ in \langmono.

% \ur{I don't quite understand the following.}
% (Thus, the resumer, bound to the value output by $p$, is equivalent to
% the residual term defined in~\ref{ssec:dynamics})

As with stage~\bbone\ terms, the splitting algorithm for stage~\bbtwo\ terms
as specified by the
judgment $e \splittwosym [p,l.r]$ in \cref{fig:termSplitTwo}, proceeds
recursively on the structure of~$e$.  In the terminal cases of
constants and variables, splitting generates \texttt{()} precomputations (they do not involve stage~\bbone\
subcomputations), and resumers that produce the original term. For
example, the integer constant \texttt{5} splits into the
precomputation \texttt{()} and \texttt{l=>5}.
